---
title: "New Buildings"
format: html
---

## Loading Libraries + Data

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(janitor)
library(ggplot2)
```


```{r}

permits <- read_csv("data-raw/Issued_Construction_Permits_20260713.csv") 

permits |> colnames()

permits |> distinct(`Work Class`)

permits_clean <- permits |> clean_names(
) |> 
  filter(work_class == "New", permit_type_desc == "Building Permit") |> # I looked it up and multifamily housing is considered commerical!! no need to filter out
  select(permit_type, permit_num, permit_class_mapped, permit_class, work_class, condominium, project_name, description, tcad_id, issued_date,calendar_year_issued, fiscal_year_issued, status_current, expires_date, completed_date, council_district, housing_units,number_of_floors, total_job_valuation, original_zip, council_district, jurisdiction, link, project_id, master_permit_num,contractor_company_name,contractor_phone,contractor_zip, applicant_organization, applicant_phone, certificate_of_occupancy, original_address_1) |>  # KH: just grabbed the ones that seemed important
  filter(calendar_year_issued >= 1996) ## grabbing only thirty years

permits_clean 
```

## Determining Permit Class

Now, I need to grab only the permits associated with housing. I already filtered out for only new building permits. 

From the Open Data Portal. I can determine that the following clases are housing. 

COMMERICAL PERMITS (KH: Need to double check with city)
C- 101 Single Family Houses
C- 103 Two Family Bldgs
C- 104 Three & Four Family Bldgs
C- 105 Five or More Family Bldgs

RESIDENTIAL PERMITS (KH: grabbing these because housing stock doesn't always have to be commerical!)
R- 101 Single Family Houses
R- 102 Secondary Apartment
R- 103 Two Family Bldgs

Or this is all pointless and I can just say that if it has a housing unit then we consider it.

```{r}

housing_types <- c("C- 101 Single Family Houses", 
                   "C- 103 Two Family Bldgs", 
                   "C- 104 Three & Four Family Bldgs", 
                   "C- 105 Five or More Family Blgds",  # KH: Double check on this
                   "R- 101 Single Family Houses", 
                   "R- 102 Secondary Apartment", 
                   "R- 103 Two Family Bldgs")

housing_types

housing <- permits_clean |> 
  filter(permit_class %in% housing_types) 

housing |> write_rds("data-export/housing_clean.rds") # need this for later
```

## Getting rid of aborted projects

Some projects did not get off the ground. We are making the decision to only consider active and final constructions. 

```{r}
canceled_projects <- housing |> filter(!status_current == "Final") |> #5,898 rows
  filter(!status_current == "Active") |> 
  distinct(status_current) |> 
  pull()

real_houses <- 
  housing |> filter(!status_current%in%canceled_projects)
```
## Getting rid of Not housing

NOTE: This section occurred towards the end of the process. If it feels like it is ahead of itself, that is because it sorta

```{r}

multifam <- c("C- 104 Three & Four Family Bldgs", "C- 105 Five or More Family Bldgs", "C- 103 Two Family Bldgs") 

multifams <- real_houses |> 
  filter(permit_class %in% multifam)

multifams |> 
  filter(is.na(housing_units)) |>
  write_csv("data-checks/multifam_na.csv") # this one was true, only need to fix two rows

not_housing_also <- 
  multifams |> 
  filter(is.na(housing_units)) |>
  filter(!permit_num=="1999-015531 BP") |> 
  filter(!permit_num=="2000-005066 BP") |> 
  pull(tcad_id)

sf_housing <- c("C- 101 Single Family Houses",
"R- 101 Single Family Houses")

sf_housing_permits <- real_houses |> 
  filter(permit_class %in% sf_housing)

sf_housing_permits |> filter(is.na(housing_units)) |> 
  write_csv("data-checks/singlefam_na.csv")# 10,305 rows, still includes housing??
  
```

For multifamily housing, we only need to fixed five of the 23 rows. For single family its more crazy... 

After I loaded my bad single rows into sheets, I scanned through for sample keywords that signal housing. Here is the list I found: 
- "Story"
- "Sty"
- "Stry"
- "Residence"
- "Condo" 

The primary culprit seems to be pools. At first I thought we could filter out pools but we run into the garage problem again. So, after I filtered for those words, I started to systemtically guess-n-check for more words I could be missing, going through the results and pulling more words. 

Yes, I could have used AI for this but we need to be careful. After I get this approved, I will be pulling their tcads and filtering those out of our original data. (Also moving some things around so everything works currently)

```{r}
keywords <- c("Story", "Sty", "Stry", "Residence", "Condo", "W/Attached Garage", "Secondary Apartment", "Secondary Apt", "Residential", "Duplex", "New Sf Res","New Single", "New Two", "manufactured", "Manufacutred", "New Group Home W/5 Bedrooms", "New One", "Sf Res. W/Attched Garage Covered Porch & Uncovered Patio")

not_housing <- sf_housing_permits |> filter(is.na(housing_units)) |> 
  filter(!str_detect(description, regex(str_c(keywords, collapse = "|"), ignore_case = TRUE))) |> pull(tcad_id)

really_real_houses <- real_houses |> 
  filter(!tcad_id%in%not_housing) |> 
  filter(!tcad_id%in%not_housing_also)
```

## Overview

### Housing Permits Issued per year 

```{r}
per_year <- really_real_houses |> 
  group_by(calendar_year_issued) |> # Travis prefers calendar year over fiscal year
  count(permit_type) |> 
  filter(!calendar_year_issued==2026)

per_year |> write_csv("4data-wrapper/permits_per_year.csv")
# KH: dropping 2026 since this year isn't done
```

#### Chart 

```{r}
# KH: not cleaning this up because I don't feel like it at this moment
 per_year |> ggplot(
   aes(
     x = calendar_year_issued, 
     y = n
   )
 ) + geom_point() +
  geom_line() + 
  labs(
    title = "Housing Permits Issued per year"
  )
```
### Getting Duration

After speaking with Robbie Searcy from Austin Development Services, we determined that the best way to find how long a project is we would need to compare the "issued date" with the "completed date." However, there is one note that the certificate of occupancy "is provided to the applicant after the construction work has been completed and passed all required inspections." But this column in the data set does not have a date associated with it, so I will just heed Searcy's advice here. 

```{r}
construction_time <- really_real_houses |> 
  select(project_name, issued_date, completed_date, jurisdiction, original_address_1, applicant_organization, project_id, calendar_year_issued, permit_class, council_district, original_zip, housing_units) |> 
  mutate(project_length = 
           as.numeric(difftime(completed_date, issued_date, units = "weeks"))) |> 
  filter(project_length >= 1) # drops NAs

construction_time

construction_time |> write_rds("data-export/project_length.rds")
```

#### Averages of Completed Projects per year (in weeks) (overall)

```{r}
construction_time |> 
  filter(!calendar_year_issued==2026) |> 
  group_by(calendar_year_issued) |> 
  summarize("avg_time" = mean(project_length)) |> 
  ggplot(aes(
    x = calendar_year_issued, 
    y = avg_time
  )) + geom_point() + geom_line() + geom_vline(xintercept = 2020, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + # KH: felt important to have markers of the recession + pandemic
  labs(
    y = "Project Length in Weeks", 
    x = "Year"
  ) + 
  geom_hline(yintercept = 52, color = "blue", linetype = "dashed") # KH: Overlaying permits + duration of construction


## Average construction time has fallen from pandemic levels? I think type would be good here
   
```

#### Average Projects Lengths by permit classificiation

This includes various types of permits. Let's get a look at numbers per permit type. 

Other items I want: 
- Distinct by type
- Get multi-use specific data

```{r}
construction_time |> 
  filter(!calendar_year_issued==2026) |>
  filter(project_length >= 1) |> 
  group_by(calendar_year_issued, permit_class) |> 
  summarize("avg_time" = mean(project_length)) |> # print()

#KH: I like seeing the charts
ggplot(aes(
    x = calendar_year_issued, 
    y = avg_time, 
    group = permit_class,
    color = permit_class
  )) + 
  geom_point() + geom_jitter(width = 0.1, height = 0.1) 

#KH: High rises take longer, who could have guessed?
```
## Permits issued by District 

One of the main questions is where are most of the building permits getting issued? For this, I will do one within our 30 year window, a 20 year window and then a 10 year window, categorizing by district. Then, do individualized looks at those districts. 

### Overall view

```{r}
cols <- c("1"="#E69F00","2"="#56B4E9","3"="#009E73","4"="#F0E442","5"="#0072B2",
          "6"="#D55E00","7"="#CC79A7","8"="#000000","9"="red4","10"="#44AA99")

districts <- really_real_houses |>
  mutate(council_district = as.factor(council_district)) |> 
  group_by(calendar_year_issued, council_district) |>
  count(permit_type) 

districts |>
  ggplot(aes(x = calendar_year_issued, y = n, fill = council_district)) +
  geom_col() +
  scale_fill_manual(values = cols) + 
  geom_vline(xintercept = 2020, color = "red", linetype ="dashed") + 
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + 
  labs(
    x = "Year", 
    y = "New Construction Permits Issued"
  )
# KH: I know this looks a mess, I just needed a chart

#KH: District's 1 and 2 consistently outpace housing permits, with other districts getting minor growth in comparison.

# GET INTO MAP
```
### Table 

```{r}

district_concentration <- districts |>
  group_by(council_district) |>
  summarise(total = sum(n)) |> 
  arrange(desc(total))|> 
  print()

districts |> filter(council_district == "1")

# KH: District 1 takes the cake on permits issuesd at 16284 construction permits. D2 15486
```

### District 1

```{r}
districts |> 
  filter(council_district == "1") |>
  ggplot(aes(
    x = calendar_year_issued, 
    y = n
  )) + geom_point() + geom_line() + geom_vline(xintercept = 2020, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + # KH: felt important to have markers of the recession + pandemic
  labs(
    title = "District 1",
    y = "New Construction Permits", 
    x = "Year"
  )
```

### District 2
```{r}
districts |> 
  filter(council_district == "2") |>
  ggplot(aes(
    x = calendar_year_issued, 
    y = n
  )) + geom_point() + geom_line() + geom_vline(xintercept = 2020, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + # KH: felt important to have markers of the recession + pandemic
  labs(
    title = "District 2",
    y = "New Construction Permits", 
    x = "Year"
  )
```
### District 3
```{r}
districts |> 
  filter(council_district == "3") |>
  ggplot(aes(
    x = calendar_year_issued, 
    y = n
  )) + geom_point() + geom_line() + geom_vline(xintercept = 2020, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + # KH: felt important to have markers of the recession + pandemic
  labs(
    title = "District 3",
    y = "New Construction Permits", 
    x = "Year"
  )
```

### District 4
```{r}
districts |> 
  filter(council_district == "4") |>
  ggplot(aes(
    x = calendar_year_issued, 
    y = n
  )) + geom_point() + geom_line() + geom_vline(xintercept = 2020, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + # KH: felt important to have markers of the recession + pandemic
  labs(
    title = "District 4",
    y = "New Construction Permits", 
    x = "Year"
  )
```

### District 5
```{r}
districts |> 
  filter(council_district == "5") |>
  ggplot(aes(
    x = calendar_year_issued, 
    y = n
  )) + geom_point() + geom_line() + geom_vline(xintercept = 2020, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + # KH: felt important to have markers of the recession + pandemic
  labs(
    title = "District 5",
    y = "New Construction Permits", 
    x = "Year"
  )

# KH: They just run out of room?
```

### District 6
```{r}
districts |> 
  filter(council_district == "6") |>
  ggplot(aes(
    x = calendar_year_issued, 
    y = n
  )) + geom_point() + geom_line() + geom_vline(xintercept = 2020, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + # KH: felt important to have markers of the recession + pandemic
  labs(
    title = "District 6",
    y = "New Construction Permits", 
    x = "Year"
  )

# KH: Mostly single family homes is this area? Lots of traction before the recession, crawling back up and then drops at the Pandemic?
```

### District 7
```{r}
districts |> 
  filter(council_district == "7") |>
  ggplot(aes(
    x = calendar_year_issued, 
    y = n
  )) + geom_point() + geom_line() + geom_vline(xintercept = 2020, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + # KH: felt important to have markers of the recession + pandemic
  labs(
    title = "District 7",
    y = "New Construction Permits", 
    x = "Year"
  )

districts |> filter(council_district == "7") |> 
  arrange(desc(n))
```
### District 8
```{r}
districts |> 
  filter(council_district == "8") |>
  ggplot(aes(
    x = calendar_year_issued, 
    y = n
  )) + geom_point() + geom_line() + geom_vline(xintercept = 2020, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + # KH: felt important to have markers of the recession + pandemic
  labs(
    title = "District 8",
    y = "New Construction Permits", 
    x = "Year"
  )
```

### District 9
```{r}
districts |> 
  filter(council_district == "9") |>
  ggplot(aes(
    x = calendar_year_issued, 
    y = n
  )) + geom_point() + geom_line() + geom_vline(xintercept = 2020, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + # KH: felt important to have markers of the recession + pandemic
  geom_vline(xintercept = 2004, color = "purple", linetype ="dashed") + # KH: This is when the University Neighborhood Overlay was introduced, which wanted to foster more higher density growth in West Campus
  labs(
    title = "District 9",
    y = "New Construction Permits", 
    x = "Year"
  )

#KH: For this section, I would also want to look at renovations? Assuming that lower density structures would need to be demlioshed in order to place higher density builds would mean we need new construction permits, right?

```

### District 10
```{r}
districts |> 
  filter(council_district == "10") |>
  ggplot(aes(
    x = calendar_year_issued, 
    y = n
  )) + geom_point() + geom_line() + geom_vline(xintercept = 2020, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + # KH: felt important to have markers of the recession + pandemic
  labs(
    title = "District 10",
    y = "New Construction Permits", 
    x = "Year"
  )
```


### Project Duration per district

```{r}
time_district <- construction_time |>
  filter(!calendar_year_issued==2026) |>
  filter(project_length >= 0) |>
  mutate(council_district = as.character(council_district)) |>
  group_by(calendar_year_issued, council_district) |>
  summarize("avg_time" = mean(project_length), .groups = "drop")
  
time_district
```


```{r}
time_district |>  
ggplot(aes(
    x = calendar_year_issued, 
    y = avg_time, 
    group = council_district,
    color = council_district
  )) + 
  geom_point() + geom_line() + 
  scale_color_manual(values = cols) + 
  geom_hline(yintercept = 52, color = "red", linetype = "dashed") + 
  geom_hline(yintercept = 104, color = "purple", linetype = "dashed")

# Light blue (District 2), Orange (District 1) seems to be the quickest turnaround, could add to point of building as fast as we can and then what?

# GET INTO MAP
  
```
(KH: Do I need to do like a per capita thing? Like a ratio of total projects/length of time? I feel like average covers this.)

### Districts 1, 2, 9, 5 

Pulling out these districts since 1 & 2 are top performers, downtown and then Universtiy Area

```{r}
time_district |> 
  filter(council_district == "1") |> 
  ggplot(aes(
    x = calendar_year_issued, 
    y = avg_time
  )) + 
  geom_point() + geom_line() + labs(
    title = "Average construction time in District 1 (in weeks)"
  ) + 
  geom_hline(yintercept = 52, color = "red", linetype = "dashed") # one year mark
  
# KH: crazy growth in oughts
```
```{r}
time_district |> 
  filter(council_district == "2") |> 
  ggplot(aes(
    x = calendar_year_issued, 
    y = avg_time
  )) + 
  geom_point() + geom_line() + labs(
    title = "Average construction time in District 2 (in weeks)"
  )+ 
  geom_hline(yintercept = 52, color = "red", linetype = "dashed")
# KH: Districts 1 & 2 show similar patterns, main districts people are moving too?
```

```{r}
time_district |> 
  filter(council_district == "9") |> 
  ggplot(aes(
    x = calendar_year_issued, 
    y = avg_time
  )) + 
  geom_point() + geom_line() + labs(
    title = "Average construction time in District 9 (in weeks)"
  ) + 
  geom_hline(yintercept = 52, color = "red", linetype = "dashed")

#KH: Did the university overlay even do anything?? We should track student growth with this?
```

```{r}
time_district |> 
  filter(council_district == "5") |> 
  ggplot(aes(
    x = calendar_year_issued, 
    y = avg_time
  )) + 
  geom_point() + geom_line() + labs(
    title = "Average construction time in District 5 (in weeks)"
  )+ 
  geom_hline(yintercept = 52, color = "red", linetype = "dashed")

```
## Takeaways so far

After speaking with Travis, we came to two conclusions so far: 
- Growth is primarily concentrated in the East Side, giving light to a historical disparity between Austin's west and east sides. 
  - Districts not only getting more permits but buildings are going up faster
- In general, after more than 20 years of growing development, including a spike around the COVID-19 pandemic, Austin's developments slows to a "screeching halt." 

## ZIP

ZIP codes get a little more specific than districts. Getting general ideas from by district, but could get more detailed look here. There's 81 zip codes, so none of my fun charts. I need to get my big data pants on and learn leaflet.

```{r}

#KH: number of permits issued per year
zips_permits <- 
  really_real_houses |> 
  group_by(calendar_year_issued, original_zip) |> 
  count(permit_type) # |> print() 

#KH: project duration by zip 
zips_duration <- 
  construction_time |> 
  filter(!calendar_year_issued==2026) |>
  filter(project_length >= 1) |>
  group_by(original_zip) |> #group_by(calendar_year_issued, original_zip) if you want a time series
  summarize("avg_time (wks)" = mean(project_length), .groups = "drop") # |> print()

zips_duration |> write_rds("data-export/zip_project_duration.rds") # bringing this for datawrapper

#KH: concentration (IE, total permits issued overall)
zips_concentration <- zips_permits |> 
  group_by(original_zip) |> 
  summarize("total" = sum(n)) # |> print()

zips_concentration |> write_rds("data-export/zips_concentration.rds")
# KH: >> this into the maps

#KH: Forget to do this one upstairs, whoops: 
district_duration <- 
  construction_time |> 
  filter(!calendar_year_issued==2026) |>
  filter(project_length >= 0) |> 
  group_by(council_district) |> 
  summarize("avg_time" = mean(project_length)) 
  
```
### Top developing ZIP codes for housing

#### Since 1996

```{r}
zips_concentration |>
  filter(total >= 3000) |> 
  arrange(desc(total)) |> 
  print()
```
In the last thirty years, the city of Austin sprawled, with a majority of the ZIP codes in the top development hot spots being in Austin's subrubs, with the expection of 78723 (Mueller) and 78702 (Riverside). 

#### Since 2006 + So on

```{r}
since_2006 <-
zips_permits |> 
  group_by(original_zip) |> 
  filter(calendar_year_issued >= 2006) |> 
  summarize("total" = sum(n)) |> 
  arrange(desc(total)) |> 
  head(10) |> 

mutate("timeframe" = "from_96")

since_2016 <- 
 zips_permits |> 
  group_by(original_zip) |> 
  filter(calendar_year_issued >= 2016) |> 
  summarize("total" = sum(n)) |> 
  arrange(desc(total)) |> 
  head(10) |> 

mutate("timeframe" = "from_16")

since_2019 <-
  zips_permits |> 
  group_by(original_zip) |> 
  filter(calendar_year_issued >= 2019) |> 
  summarize("total" = sum(n)) |> 
  arrange(desc(total)) |> 
  head(10) |> 
mutate("timeframe" = "from_19")

bind_rows(since_2006, since_2016, since_2019) |> filter()

```

## Multi-family Housing

NOTE: DATA FRAME FOR THE FOLLOWING TWO SECTIONS WERE CREATED IN "Getting rid of Not housing" PLEASE ADVISE

C- 104 Three & Four Family Bldgs
C- 105 Five or More Family Bldgs


1. Count number of highrises given out in the last 30 years
2. Track where, district and zipcode, are they getting built. 
3. Track how long its taking to get them up. 

```{r}
multifams |> 
  summarise("total_units" = sum(as.numeric(housing_units), na.rm = TRUE))
```

Since 1996, there has been 1,223 permits for multifamily developments with more than housing units per project in the city of Austin, representing about 5,367 housing units.

### Permits Issued per Year

```{r}
multi_times <- multifams |> 
  group_by(calendar_year_issued) |>
  count(permit_type)

multi_times |> 
  arrange(desc(n))

multi_times|>
  ggplot(
    aes(
      x = calendar_year_issued, 
      y = n
    )
  ) + geom_point() + geom_line()
# KH: Significant drops? 
```
### Project Length

```{r}

MF_project_length <- construction_time |> 
  filter(permit_class %in% multifam) |>
  filter(!calendar_year_issued==2026) |>
  filter(project_length >= 0)
# KH: per year

 MF_project_length |> 
  group_by(calendar_year_issued) |> 
  summarize("avg_time" = mean(project_length)) |> #print() 
  ggplot(
    aes(
      x = calendar_year_issued, 
      y = avg_time
    )
  ) + 
  geom_point() + 
  geom_line() + 
  geom_hline(yintercept = 52, color = "red", linetype = "dashed") +
  geom_hline(yintercept = 104, color = "orange", linetype = "dashed") + 
  geom_hline(yintercept = 156, color = "red4", linetype = "dashed")
```

```{r}
# KH: Overall
MF_project_length |> 
  summarize("avg_time" = mean(project_length)) |> 
  print()

MF_project_length |> 
  summarize("median_time" = median(project_length)) |> 
  print()

# takes about 85.8 weeks on average to get a highrise up, about 63.4 weeks is the middle number or another kind of average.
```

### By district, ZIP

#### By district

```{r}
multifams |> 
  group_by(council_district) |> 
  count(permit_type) |> 
  arrange(desc(n))
```

```{r}
multifams |> 
  group_by(original_zip) |> 
  count(permit_type) |> 
  arrange(desc(n)) |> 
  print()
```

## Single Family housing

From our list above here are only the single family housing. But first, I want to look at how much of each of these subtypes make up of the whole. Do to that, I will be counting (again) each subtype by year. 

C- 101 Single Family Houses
R- 101 Single Family Houses

### Overall

```{r}
# Austin's housing stock by type 
by_type <- really_real_houses |> 
  group_by(permit_class) |> 
  count(permit_type)  |> 
  arrange(desc(n))

by_type |> write_csv("4data-wrapper/housing_by_type.csv")


by_type|> # print() 
  ggplot(aes(
    x = "", 
    y = n, 
    fill = permit_class
  )) + 
  geom_bar(stat = "identity", width = 0.5, color = "white") +
  coord_polar("y", start = 0) + 
  theme_void() + 
  labs(
    title = "Austin's last 30 years in housing",
    fill = "Housing Type"
  )
```

Okay, residential single family housing represents a majority of what was built in Austin over the last 30 years. And, it's not even close -- 86,371 permits.

```{r}
# Single family housing by year
sf_housing_permits |>
  group_by(calendar_year_issued, permit_class) |> 
  count(permit_type) |> # print()
  ggplot(
    aes(
      x = calendar_year_issued, 
      y = n,
      color = permit_class
    )
  ) + 
  geom_point() + 
  geom_line() + 
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + 
  geom_vline(xintercept = 2020, color = "red", linetype = "dashed")

# Single family development is trending downward. Remember 2026 is only halfway over, so there could be signs of a post-pandemic frenzy recovery. 
```

### By District and ZIP

```{r}
sf_housing_permits |> 
  group_by(council_district) |> 
  count(permit_type) |> 
  arrange(desc(n)) 
```

```{r}
sf_housing_permits |> 
  group_by(original_zip) |> 
  count(permit_type) |> 
  arrange(desc(n))
```

## Misc Findings

### Finding addresses for Photo

- Should we get within the last five/six years and make the maps for concentration since the pandemic?

```{r}
housing |> filter(original_zip==78748) |> 
  arrange(desc(calendar_year_issued)) |> 
  filter(status_current=="Final")
```

### Top Contractors (just curious)

```{r}
housing |> group_by(contractor_company_name) |> 
  count(contractor_company_name) |> 
  arrange(desc(n))|> 
  head(21) |> 
  write_csv("data-export/top_contractors.csv")

```

### Average Multifamily unit number

```{r}
multifams |> 
  count(permit_type)
# 1223 projects

multifams |> drop_na(housing_units) |> # dropped 23 projects
  summarise("avg" = mean(housing_units)) # using average bc I want to weigh outliers

```


### East Riverside

[Going through major developments](https://www.statesman.com/news/article/austin-east-riverside-light-rail-21333956.php), zipcode 78741 --> checking their housing

```{r}
real_houses |> 
  filter(original_zip == 78741) |> 
  group_by(calendar_year_issued) |> 
  count(permit_type) |> 
  ggplot(aes(
    x = calendar_year_issued, 
    y = n
  )) + geom_point() + geom_line() + 
  labs(
    title = "Building permits issued in East Riverside"
  )
```