---
title: "New Buildings"
format: html
---

## Loading Libraries + Data

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(janitor)
library(ggplot2)
```


```{r}

permits <- read_csv("data-raw/Issued_Construction_Permits_20260604.csv") 

permits |> colnames()

permits_clean <- permits |> clean_names(
) |> 
  filter(work_class == "New", permit_type_desc == "Building Permit") |> # I looked it up and multifamily housing is considered commerical!! no need to filter out
  select(permit_type, permit_num, permit_class_mapped, permit_class, work_class, condominium, project_name, description, tcad_id, issued_date,calendar_year_issued, fiscal_year_issued, status_current, expires_date, completed_date, council_district, housing_units,number_of_floors, total_job_valuation, original_zip, council_district, jurisdiction, link, project_id, master_permit_num,contractor_company_name,contractor_phone,contractor_zip, applicant_organization, applicant_phone, certificate_of_occupancy, original_address_1) |>  # KH: just grabbed the ones that seemed important
  filter(calendar_year_issued >= 1996) ## grabbing only thirty years

permits_clean
```

## Determining Permit Class

Now, I need to grab only the permits associated with housing. I already filtered out for only new building permits. 

From the Open Data Portal. I can determine that the following clases are housing. 

COMMERICAL PERMITS (KH: Need to double check with city)
C- 101 Single Family Houses
C- 103 Two Family Bldgs
C- 104 Three & Four Family Bldgs
C- 105 Five or More Family Bldgs

RESIDENTIAL PERMITS (KH: grabbing these because housing stock doesn't always have to be commerical!)
R- 101 Single Family Houses
R- 102 Secondary Apartment
R- 103 Two Family Bldgs
R- 436 Addn to increase housing units

Or this is all pointless and I can just say that if it has a housing unit then we consider it.

```{r}

housing_types <- c("C- 101 Single Family Houses", 
                   "C- 103 Two Family Bldgs", 
                   "C- 104 Three & Four Family Bldgs", 
                   "C- 105 Five or More Family Blgds", 
                   "R- 101 Single Family Houses", 
                   "R- 102 Secondary Apartment", 
                   "R- 103 Two Family Bldgs")

housing_types

housing <- permits_clean |> 
  filter(permit_class %in% housing_types) 

housing |> write_rds("data-export/housing_clean.rds") # need this for later

housing |> distinct(permit_class)
```

## Overview

### Housing Permits Issued per year 

```{r}
per_year <- housing |> 
  group_by(calendar_year_issued) |> # Travis prefers calendar year over fiscal year
  count(permit_type) |> 
  filter(!calendar_year_issued==2026)
# KH: dropping 2026 since this year isn't done
```

#### Chart 

```{r}
# KH: not cleaning this up because I don't feel like it at this moment
 per_year |> ggplot(
   aes(
     x = calendar_year_issued, 
     y = n
   )
 ) + geom_point() +
  geom_line() + 
  labs(
    title = "Housing Permits Issued per year"
  )
```
### Getting Duration

After speaking with Robbie Searcy from Austin Development Services, we determined that the best way to find how long a project is we would need to compare the "issued date" with the "completed date." However, there is one note that the certificate of occupancy "is provided to the applicant after the constructino work has been completed and passed all required inspections." But this column in the data set does not have a date associated with it, so I will just heed Searcy's advice here. 

```{r}
construction_time <- housing |> 
  select(project_name, issued_date, completed_date, jurisdiction, original_address_1, applicant_organization, project_id, calendar_year_issued, permit_class, council_district, original_zip, housing_units) |> 
  mutate(project_length = 
           as.numeric(difftime(completed_date, issued_date, units = "weeks"))) |> 
  filter(project_length >= 1) # drops NAs

construction_time

construction_time |> write_rds("data-export/project_length.rds")
```

#### Averages of Completed Projects per year (in weeks) (overall)

```{r}
construction_time |> 
  filter(!calendar_year_issued==2026) |> 
  group_by(calendar_year_issued) |> 
  summarize("avg_time" = mean(project_length)) |> 
  ggplot(aes(
    x = calendar_year_issued, 
    y = avg_time
  )) + geom_point() + geom_line() + geom_vline(xintercept = 2020, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + # KH: felt important to have markers of the recession + pandemic
  labs(
    y = "Project Length in Weeks", 
    x = "Year"
  )
## Average construction time has fallen from pandemic levels? I think type would be good here
   
```

#### Average Projects Lengths by permit classificiation

This includes various types of permits. Let's get a look at numbers per permit type. 

Other items I want: 
- Distinct by type
- Get multi-use specific data

```{r}
construction_time |> 
  filter(!calendar_year_issued==2026) |>
  filter(project_length >= 1) |> 
  group_by(calendar_year_issued, permit_class) |> 
  summarize("avg_time" = mean(project_length)) |> # print()

#KH: I like seeing the charts
ggplot(aes(
    x = calendar_year_issued, 
    y = avg_time, 
    group = permit_class,
    color = permit_class
  )) + 
  geom_point() + geom_jitter(width = 0.1, height = 0.1) 

#KH: High rises take longer, who could have guessed?
```
## Permits issued by District 

One of the main questions is where are most of the building permits getting issued? For this, I will do one within our 30 year window, a 20 year window and then a 10 year window, categorizing by district. Then, do individualized looks at those districts. 

### Overall view

```{r}
cols <- c("1"="#E69F00","2"="#56B4E9","3"="#009E73","4"="#F0E442","5"="#0072B2",
          "6"="#D55E00","7"="#CC79A7","8"="#000000","9"="red4","10"="#44AA99")

districts <- housing |>
  mutate(council_district = as.factor(council_district)) |> 
  group_by(calendar_year_issued, council_district) |>
  count(permit_type) 

districts |>
  ggplot(aes(x = calendar_year_issued, y = n, fill = council_district)) +
  geom_col() +
  scale_fill_manual(values = cols) + 
  geom_vline(xintercept = 2020, color = "red", linetype ="dashed") + 
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + 
  labs(
    x = "Year", 
    y = "New Construction Permits Issued"
  )
# KH: I know this looks a mess, I just needed a chart

#KH: District's 1 and 2 consistently outpace housing permits, with other districts getting minor growth in comparison.

# GET INTO MAP
```
### Table 

```{r}

district_concentration <- districts |>
  group_by(council_district) |>
  summarise(total = sum(n)) |> 
  print()

districts |> filter(council_district == "1")

# KH: District 1 takes the cake on permits issuesd at 17,283 construction permits. 
```

### District 1

```{r}
districts |> 
  filter(council_district == "1") |>
  ggplot(aes(
    x = calendar_year_issued, 
    y = n
  )) + geom_point() + geom_line() + geom_vline(xintercept = 2020, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + # KH: felt important to have markers of the recession + pandemic
  labs(
    title = "District 1",
    y = "New Construction Permits", 
    x = "Year"
  )
```

### District 2
```{r}
districts |> 
  filter(council_district == "2") |>
  ggplot(aes(
    x = calendar_year_issued, 
    y = n
  )) + geom_point() + geom_line() + geom_vline(xintercept = 2020, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + # KH: felt important to have markers of the recession + pandemic
  labs(
    title = "District 2",
    y = "New Construction Permits", 
    x = "Year"
  )
```
### District 3
```{r}
districts |> 
  filter(council_district == "3") |>
  ggplot(aes(
    x = calendar_year_issued, 
    y = n
  )) + geom_point() + geom_line() + geom_vline(xintercept = 2020, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + # KH: felt important to have markers of the recession + pandemic
  labs(
    title = "District 3",
    y = "New Construction Permits", 
    x = "Year"
  )
```

### District 4
```{r}
districts |> 
  filter(council_district == "4") |>
  ggplot(aes(
    x = calendar_year_issued, 
    y = n
  )) + geom_point() + geom_line() + geom_vline(xintercept = 2020, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + # KH: felt important to have markers of the recession + pandemic
  labs(
    title = "District 4",
    y = "New Construction Permits", 
    x = "Year"
  )
```

### District 5
```{r}
districts |> 
  filter(council_district == "5") |>
  ggplot(aes(
    x = calendar_year_issued, 
    y = n
  )) + geom_point() + geom_line() + geom_vline(xintercept = 2020, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + # KH: felt important to have markers of the recession + pandemic
  labs(
    title = "District 5",
    y = "New Construction Permits", 
    x = "Year"
  )

# KH: They just run out of room?
```

### District 6
```{r}
districts |> 
  filter(council_district == "6") |>
  ggplot(aes(
    x = calendar_year_issued, 
    y = n
  )) + geom_point() + geom_line() + geom_vline(xintercept = 2020, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + # KH: felt important to have markers of the recession + pandemic
  labs(
    title = "District 6",
    y = "New Construction Permits", 
    x = "Year"
  )

# KH: Mostly single family homes is this area? Lots of traction before the recession, crawling back up and then drops at the Pandemic?
```

### District 7
```{r}
districts |> 
  filter(council_district == "7") |>
  ggplot(aes(
    x = calendar_year_issued, 
    y = n
  )) + geom_point() + geom_line() + geom_vline(xintercept = 2020, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + # KH: felt important to have markers of the recession + pandemic
  labs(
    title = "District 7",
    y = "New Construction Permits", 
    x = "Year"
  )
```
### District 8
```{r}
districts |> 
  filter(council_district == "8") |>
  ggplot(aes(
    x = calendar_year_issued, 
    y = n
  )) + geom_point() + geom_line() + geom_vline(xintercept = 2020, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + # KH: felt important to have markers of the recession + pandemic
  labs(
    title = "District 8",
    y = "New Construction Permits", 
    x = "Year"
  )
```

### District 9
```{r}
districts |> 
  filter(council_district == "9") |>
  ggplot(aes(
    x = calendar_year_issued, 
    y = n
  )) + geom_point() + geom_line() + geom_vline(xintercept = 2020, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + # KH: felt important to have markers of the recession + pandemic
  geom_vline(xintercept = 2004, color = "purple", linetype ="dashed") + # KH: This is when the University Neighborhood Overlay was introduced, which wanted to foster more higher density growth in West Campus
  labs(
    title = "District 9",
    y = "New Construction Permits", 
    x = "Year"
  )

#KH: For this section, I would also want to look at renovations? Assuming that lower density structures would need to be demlioshed in order to place higher density builds would mean we need new construction permits, right?

```

### District 10
```{r}
districts |> 
  filter(council_district == "10") |>
  ggplot(aes(
    x = calendar_year_issued, 
    y = n
  )) + geom_point() + geom_line() + geom_vline(xintercept = 2020, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + # KH: felt important to have markers of the recession + pandemic
  labs(
    title = "District 10",
    y = "New Construction Permits", 
    x = "Year"
  )
```


### Project Duration per district

```{r}
time_district <- construction_time |> 
  filter(!calendar_year_issued==2026) |>
  filter(project_length >= 0 |>
  mutate(council_district = as.character(council_district)) |>
  group_by(calendar_year_issued, council_district) |>
  summarize("avg_time" = mean(project_length), .groups = "drop")

# time_distrcit |> print()

time_district |>  
ggplot(aes(
    x = calendar_year_issued, 
    y = avg_time, 
    group = council_district,
    color = council_district
  )) + 
  geom_point() + geom_line() + 
  scale_color_manual(values = cols) + 
  geom_hline(yintercept = 52, color = "red", linetype = "dashed") + 
  geom_hline(yintercept = 104, color = "purple", linetype = "dashed")

# Light blue (District 2), Orange (District 1) seems to be the quickest turnaround, could add to point of building as fast as we can and then what?

# GET INTO MAP
  
```
(KH: Do I need to do like a per capita thing? Like a ratio of total projects/length of time? I feel like average covers this.)

### Districts 1, 2, 9, 5 

Pulling out these districts since 1 & 2 are top performers, downtown and then Universtiy Area

```{r}
time_district |> 
  filter(council_district == "1") |> 
  ggplot(aes(
    x = calendar_year_issued, 
    y = avg_time
  )) + 
  geom_point() + geom_line() + labs(
    title = "Average construction time in District 1 (in weeks)"
  ) + 
  geom_hline(yintercept = 52, color = "red", linetype = "dashed") # one year mark
  
# KH: crazy growth in oughts
```
```{r}
time_district |> 
  filter(council_district == "2") |> 
  ggplot(aes(
    x = calendar_year_issued, 
    y = avg_time
  )) + 
  geom_point() + geom_line() + labs(
    title = "Average construction time in District 2 (in weeks)"
  )+ 
  geom_hline(yintercept = 52, color = "red", linetype = "dashed")
# KH: Districts 1 & 2 show similar patterns, main districts people are moving too?
```

```{r}
time_district |> 
  filter(council_district == "9") |> 
  ggplot(aes(
    x = calendar_year_issued, 
    y = avg_time
  )) + 
  geom_point() + geom_line() + labs(
    title = "Average construction time in District 9 (in weeks)"
  ) + 
  geom_hline(yintercept = 52, color = "red", linetype = "dashed")

#KH: Did the university overlay even do anything?? We should track student growth with this?
```

```{r}
time_district |> 
  filter(council_district == "5") |> 
  ggplot(aes(
    x = calendar_year_issued, 
    y = avg_time
  )) + 
  geom_point() + geom_line() + labs(
    title = "Average construction time in District 5 (in weeks)"
  )+ 
  geom_hline(yintercept = 52, color = "red", linetype = "dashed")

```
## Takeaways so far

After speaking with Travis, we came to two conclusions so far: 
- Growth is primarily concentrated in the East Side, giving light to a historical disparity between Austin's west and east sides. 
  - Districts not only getting more permits but buildings are going up faster
- In general, after more than 20 years of growing development, including a spike around the COVID-19 pandemic, Austin's developments slows to a "screeching halt." 

## ZIP

ZIP codes get a little more specific than districts. Getting general ideas from by district, but could get more detailed look here. There's 81 zip codes, so none of my fun charts. I need to get my big data pants on and learn leaflet.

```{r}

#KH: number of permits issued per year
zips_permits <- 
  housing |> 
  group_by(calendar_year_issued, original_zip) |> 
  count(permit_type) # |> print()

#KH: project duration by zip 
zips_duration <- 
  construction_time |> 
  filter(!calendar_year_issued==2026) |>
  filter(project_length >= 1) |>
  group_by(original_zip) |> #group_by(calendar_year_issued, original_zip) if you want a time series
  summarize("avg_time (wks)" = mean(project_length), .groups = "drop") # |> print()


#KH: concentration (IE, total permits issued overall)
zips_concentration <- zips_permits |> 
  group_by(original_zip) |> 
  summarize("total" = sum(n)) # |> print()
# KH: >> this into the maps

#KH: Forget to do this one upstairs, whoops: 
district_duration <- 
  construction_time |> 
  filter(!calendar_year_issued==2026) |>
  filter(project_length >= 0) |> 
  group_by(council_district) |> 
  summarize("avg_time" = mean(project_length))
  
```
### Top developing ZIP codes for housing

#### Since 1996

```{r}
zips_concentration |> filter(total >= 3600) |> 
arrange(desc(total)) |> print()
```
In the last thirty years, the city of Austin sprawled, with a majority of the ZIP codes in the top development hot spots being in Austin's subrubs, with the expection of 78723 (Mueller) and 78702 (Riverside). 

#### Since 2006 + So on

```{r}
since_2006 <-
zips_permits |> 
  group_by(original_zip) |> 
  filter(calendar_year_issued >= 2006) |> 
  summarize("total" = sum(n)) |> 
  arrange(desc(total)) |> 
  head(10) |> 
  print()
# mutate("timeframe" = "from_96")

since_2016 <- 
 zips_permits |> 
  group_by(original_zip) |> 
  filter(calendar_year_issued >= 2016) |> 
  summarize("total" = sum(n)) |> 
  arrange(desc(total)) |> 
  head(10) |> 
  print() 
# mutate("timeframe" = "from_16")

since_2019 <-
  zips_permits |> 
  group_by(original_zip) |> 
  filter(calendar_year_issued >= 2019) |> 
  summarize("total" = sum(n)) |> 
  arrange(desc(total)) |> 
  head(10) |> 
  print()
# mutate("timeframe" = "from_19")

# bind_rows(since_2006, since_2016, since_2019) |> distinct(original_zip)

```


### East Riverside

[Going through major developments](https://www.statesman.com/news/article/austin-east-riverside-light-rail-21333956.php), zipcode 78741 --> checking their housing

```{r}
housing |> 
  filter(original_zip == 78741) |> 
  group_by(calendar_year_issued) |> 
  count(permit_type) |> 
  ggplot(aes(
    x = calendar_year_issued, 
    y = n
  )) + geom_point() + geom_line() + 
  labs(
    title = "Building permits issued in East Riverside"
  )
```

## Multi-family Housing

C- 104 Three & Four Family Bldgs
C- 105 Five or More Family Bldgs


1. Count number of highrises given out in the last 30 years
2. Track where, district and zipcode, are they getting built. 
3. Track how long its taking to get them up. 

```{r}

multifam <- c("C- 104 Three & Four Family Bldgs", "C- 105 Five or More Family Bldgs") 

multifams <- housing |> 
  filter(permit_class %in% multifam)

multifams |> 
  summarise("total_units" = sum(as.numeric(housing_units), na.rm = TRUE))
```

Since 1996, there has been 1,161 permits for multifamily developments with more than 5 housing units per project in the city of Austin, representing about 5,377 housing units.

### Permits Issued per Year

```{r}
multifams |> 
  group_by(calendar_year_issued) |>
  count(permit_type) |>
  ggplot(
    aes(
      x = calendar_year_issued, 
      y = n
    )
  ) + geom_point() + geom_line()
# KH: Significant drops? 
```
### Project Length

```{r}

MF_project_length <- construction_time |> 
  filter(permit_class %in% multifam) |>
  filter(!calendar_year_issued==2026) |>
  filter(project_length >= 0)
# KH: per year

 MF_project_length |> 
  group_by(calendar_year_issued) |> 
  summarize("avg_time" = mean(project_length)) |> #print() 
  ggplot(
    aes(
      x = calendar_year_issued, 
      y = avg_time
    )
  ) + 
  geom_point() + 
  geom_line() + 
  geom_hline(yintercept = 52, color = "red", linetype = "dashed") +
  geom_hline(yintercept = 104, color = "orange", linetype = "dashed") + 
  geom_hline(yintercept = 156, color = "red4", linetype = "dashed")
```

```{r}
# KH: Overall
MF_project_length |> 
  summarize("avg_time" = mean(project_length)) |> 
  print()

MF_project_length |> 
  summarize("median_time" = median(project_length)) |> 
  print()

# takes about 81.4 weeks on average to get a highrise up, about 60 weeks is the middle number or another kind of average.
```

### By district, ZIP

#### By district

```{r}
multifams |> 
  group_by(council_district) |> 
  count(permit_type) |> 
  arrange(desc(n))
```

```{r}
multifams |> 
  group_by(original_zip) |> 
  count(permit_type) |> 
  arrange(desc(n)) |> 
  print()
```

## Single Family housing

From our list above here are only the single family housing. But first, I want to look at how much of each of these subtypes make up of the whole. Do to that, I will be counting (again) each subtype by year. 

C- 101 Single Family Houses
R- 101 Single Family Houses

### Overall

```{r}
# Creating SF data frame
sf_housing <- c("C- 101 Single Family Houses",
"R- 101 Single Family Houses")

sf_housing_permits <- housing |> 
  filter(permit_class %in% sf_housing)

# Austin's housing stock by type 
housing |> 
  group_by(permit_class) |> 
  count(permit_type) |> # print() 
  ggplot(aes(
    x = "", 
    y = n, 
    fill = permit_class
  )) + 
  geom_bar(stat = "identity", width = 0.5, color = "white") +
  coord_polar("y", start = 0) + 
  theme_void() + 
  labs(
    title = "Austin's last 30 years in housing",
    fill = "Housing Type"
  )
```

Okay, residential single family housing represents a majority of what was built in Austin over the last 30 years. And, it's not even close -- 86,371 permits.

```{r}
# Single family housing by year
sf_housing_permits |>
  group_by(calendar_year_issued, permit_class) |> 
  count(permit_type) |> # print()
  ggplot(
    aes(
      x = calendar_year_issued, 
      y = n,
      color = permit_class
    )
  ) + 
  geom_point() + 
  geom_line() + 
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + 
  geom_vline(xintercept = 2020, color = "red", linetype = "dashed")

# Single family development is trending downward. Remember 2026 is only halfway over, so there could be signs of a post-pandemic frenzy recovery. 
```

### By District and ZIP

```{r}
sf_housing_permits |> 
  group_by(council_district, permit_class) |> 
  count(permit_type) |> arrange(desc(n)) 
```

```{r}
sf_housing_permits |> 
  group_by(original_zip, permit_class) |> 
  count(permit_type) |> arrange(desc(n))
```

## Further Questions

- Should we get within the last five/six years and make the maps for concentration since the pandemic?
