---
title: "1.5-FormattingforDW"
format: html
---

## Libraries and Loading data

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(janitor)
library(sf)
```

```{r}
housing_clean <- read_rds("data-export/housing_clean.rds")

zip_concentrations <- read_rds("data-export/zips_concentration.rds")

zip_lengths <- read_rds("data-export/zip_project_duration.rds")

CEN_zip_geom <- read_sf("census_geom/tl_2025_us_zcta520.shp")
```


## Goals

- Create one data frame that includes the following: 
   - Zipcode
   - Number of Housing Permits
   - Average Project Duration in that zip code
   - Percentage of particular permits
    - We are spliting this cleanly into Single Family and Multifamily
   - Zip code geometry 
- Export into GeoSPON file for the Castle template

## Creating Single Family + Multifamily flags 

I feel like the easiest way to condense the permit_class is through creating flags, which is agruably my least favorite thing to do... 
```{r}

single_family <- c("R- 101 Single Family Houses", "C- 101 Single Family Houses")
condensed_zips <- 
  housing_clean |> 
  mutate(housing_type = if_else(
    permit_class %in% single_family,
    "Single Family",
    "Multi Family"
  ))
```

## Getting percentages

```{r}
zip_class_pct <- condensed_zips |>
  group_by(original_zip, housing_type) |>
  summarize(n= n(), .groups = "drop") |>
  group_by(original_zip) |>
  mutate(
    zip_total = sum(n),
    pct = round(n / zip_total * 100, 1)
  ) |>
  ungroup() |> 
  select(!n)

zip_class_pct
```


```{r}

zip_values <- zip_class_pct |> 
  pivot_wider(
    names_from = housing_type, 
    values_from = pct,
    values_fill = 0) |> 
  clean_names() |> 
  rename("zipcode" = original_zip) |> 
  mutate("zipcode" = as.character(zipcode))
# KH: considering simplifing this into residential vs commerical percent or single v multi family housing?

zip_values
```

## Cleaning Census Data

```{r}
acl_zips <- zip_values |> 
  distinct(zipcode) |> 
  pull() # getting filter list for austin

austin_zip_geom <- CEN_zip_geom |> 
  filter(ZCTA5CE20 %in% acl_zips) |> 
  rename("zipcode" = ZCTA5CE20) # for join later 
```

I have an "NA" field in the constructon permit data. I will isolate that, filter it out + pull a few details to ensure we are not missing anything + notation purposes. 

```{r}
acl <- austin_zip_geom |> 
  distinct(zipcode) |> 
  pull()

NA_permits <- zip_values |> 
  filter(!zipcode%in%acl)

NA_permits
  
```

Permits where the ZIP was unavailable (99 total) are 98% single family homes and 100% residential.

## Joining with Geometry 

```{r}
zips_for_map <- zip_values |> 
  left_join(austin_zip_geom, by = "zipcode")

zips_for_map |> write_sf("data-export/CEN_zipcode_map.geojson")

zips_for_map |> st_write("data-export/CEN_zipcode_map_2.geojson")
```


## In progress map

First I need to handcheck all scores with an accurancy under 0.8. Filtering out for those and then exporting into a csv

```{r}
geo_housing <- read_csv("data-export/geo_housing.csv")

bad_geos <- geo_housing |> filter(geocodio_accuracy_score < 0.8)

bad_geos |> write_csv("data-export/bad_geos_housing.csv")
```

~ fixing using Google Dev Tools ~

```{r}
fixed_bad_geos <- read_csv("data-raw/fixed_bad_geos.csv")

fixed_geo_housing <- 
  bind_rows(geo_housing |> 
              filter(!geocodio_accuracy_score < 0.8), fixed_bad_geos) |> 
  arrange(desc(issued_date)) # this just to make it tidy aagin

fixed_geo_housing #5,119, still got all my rows

fixed_geo_housing |> colnames()

wip_map <- fixed_geo_housing |> 
  select(project_name, description, clean_address, geocodio_latitude, geocodio_longitude)

wip_map |> write.csv("data-export/fixed_wip_map.csv")
```

