---
title: "New Buildings"
format: html
---

## Loading Libraries + Data

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(janitor)
library(ggplot2)
```


```{r}

permits <- read_csv("data-raw/Issued_Construction_Permits_20260604.csv") 

permits |> colnames()

permits_clean <- permits |> clean_names(
) |> 
  filter(work_class == "New", permit_type_desc == "Building Permit") |> # I looked it up and multifamily housing is considered commerical!! no need to filter out
  select(permit_type, permit_num, permit_class_mapped, permit_class, work_class, condominium, project_name, description, tcad_id, issued_date,calendar_year_issued, fiscal_year_issued, status_current, expires_date, completed_date, council_district, housing_units,number_of_floors, total_job_valuation, original_zip, council_district, jurisdiction, link, project_id, master_permit_num,contractor_company_name,contractor_phone,contractor_zip, applicant_organization, applicant_phone, certificate_of_occupancy, original_address_1) |>  # KH: just grabbed the ones that seemed important
  filter(calendar_year_issued >= 1996) ## grabbing only thirty years

```

## Determining Permit Class

Now, I need to grab only the permits associated with housing. I already filtered out for only new building permits. 

From the Open Data Portal. I can determine that the following clases are housing. 

COMMERICAL PERMITS (KH: Need to double check with city)
C- 101 Single Family Houses
C- 103 Two Family Bldgs
C- 104 Three & Four Family Bldgs
C- 105 Five or More Family Bldgs
C- 106 Mixed use 

RESIDENTIAL PERMITS (KH: grabbing these because housing stock doesn't always have to be commerical!)
R- 101 Single Family Houses
R- 102 Secondary Apartment
R- 103 Two Family Bldgs
R- 436 Addn to increase housing units

Or this is all pointless and I can just say that if it has a housing unit then we consider it.

```{r}

housing_types <- c("C- 101 Single Family Houses", "C- 103 Two Family Bldgs", "C- 104 Three & Four Family Bldgs", "C- 105 Five or More Family Blgds", "C- 106 Mixed Use", "R- 101 Single Family Houses", "R- 102 Secondary Apartment", "R- 103 Two Family Bldgs", "R- 436 Addn to increase housing units")

housing_types

housing <- permits_clean |> 
  filter(permit_class %in% housing_types)
```

## Overview

### Housing Permits Issued per year 

```{r}
per_year <- housing |> group_by(calendar_year_issued) |> # Travis prefers calendar year over fiscal year
  count(permit_type) |> filter(!calendar_year_issued==2026) # KH: dropping 2026 since this year isn't done
```

#### Chart 

```{r}
# KH: not cleaning this up because I don't feel like it at this moment
 per_year |> ggplot(
   aes(
     x = calendar_year_issued, 
     y = n
   )
 ) + geom_point() +
  geom_line() + 
  labs(
    title = "Housing Permits Issued per year"
  )
```
### Getting Duration

After speaking with Robbie Searcy from Austin Development Services, we determined that the best way to find how long a project is we would need to compare the "issued date" with the "completed date." However, there is one note that the certificate of occupancy "is provided to the applicant after the constructino work has been completed and passed all required inspections." But this column in the data set does not have a date associated with it, so I will just heed Searcy's advice here. 

```{r}
construction_time <- housing |> 
  select(project_name, issued_date, completed_date, jurisdiction, original_address_1, applicant_organization, project_id, calendar_year_issued, permit_class, council_district) |> 
  mutate(project_length = 
           as.numeric(difftime(completed_date, issued_date, units = "weeks")))
construction_time
# check 5918 Lookout Mountain Dr, project length negative?? think completed and issued dates are swapped.
```

#### Averages of Completed Projects per year (in weeks) (overall)

```{r}
construction_time |> 
  filter(!calendar_year_issued==2026) |>
  filter(project_length >= 1) |> 
  group_by(calendar_year_issued) |> 
  summarize("avg_time" = mean(project_length)) |> 
  ggplot(aes(
    x = calendar_year_issued, 
    y = avg_time
  )) + geom_point() + geom_line() + geom_vline(xintercept = 2020, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + # KH: felt important to have markers of the recession + pandemic
  labs(
    y = "Project Length in Weeks", 
    x = "Year"
  )
## Average construction time has fallen from pandemic levels? I think type would be good here
   
```

#### Average Projects Lengths by permit classificiation

This includes various types of permits. Let's get a look at numbers per permit type. 

Other items I want: 
- Distinct by type
- Get multi-use specific data

```{r}
construction_time |> 
  filter(!calendar_year_issued==2026) |>
  filter(project_length >= 1) |> 
  group_by(calendar_year_issued, permit_class) |> 
  summarize("avg_time" = mean(project_length)) |> # print()

#KH: I like seeing the charts
ggplot(aes(
    x = calendar_year_issued, 
    y = avg_time, 
    group = permit_class,
    color = permit_class
  )) + 
  geom_point() + geom_jitter(width = 0.1, height = 0.1) 

#KH: High rises take longer, who could have guessed?
```
## Permits issued by District 

One of the main questions is where are most of the building permits getting issued? For this, I will do one within our 30 year window, a 20 year window and then a 10 year window, categorizing by district. Then, do individualized looks at those districts. 

### Overall view

```{r}
cols <- c("1"="#E69F00","2"="#56B4E9","3"="#009E73","4"="#F0E442","5"="#0072B2",
          "6"="#D55E00","7"="#CC79A7","8"="#000000","9"="red4","10"="#44AA99")

districts <- housing |>
  mutate(council_district = as.factor(council_district)) |>  # ← add this
  group_by(calendar_year_issued, council_district) |>
  count(permit_type) 

districts |>
  ggplot(aes(x = calendar_year_issued, y = n, fill = council_district)) +
  geom_col() +
  scale_fill_manual(values = cols) + 
  geom_vline(xintercept = 2020, color = "red", linetype ="dashed") + 
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + 
  labs(
    x = "Year", 
    y = "New Construction Permits Issued"
  )
# KH: I know this looks a mess, I just needed a chart

#KH: District's 1 and 2 consistently outpace housing permits, with other districts getting minor growth in comparison.
```
### Table 

```{r}
districts |>
  group_by(council_district) |>
  summarise(total = sum(n)) |> 
  print()

# KH: District 1 takes the cake on permits issuesd at 17,283 construction permits for housing being issues. 
```

### District 1

```{r}
districts |> 
  filter(council_district == "1") |>
  ggplot(aes(
    x = calendar_year_issued, 
    y = n
  )) + geom_point() + geom_line() + geom_vline(xintercept = 2020, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + # KH: felt important to have markers of the recession + pandemic
  labs(
    title = "District 1",
    y = "New Construction Permits", 
    x = "Year"
  )
```

### District 2
```{r}
districts |> 
  filter(council_district == "2") |>
  ggplot(aes(
    x = calendar_year_issued, 
    y = n
  )) + geom_point() + geom_line() + geom_vline(xintercept = 2020, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + # KH: felt important to have markers of the recession + pandemic
  labs(
    title = "District 2",
    y = "New Construction Permits", 
    x = "Year"
  )
```
### District 3
```{r}
districts |> 
  filter(council_district == "3") |>
  ggplot(aes(
    x = calendar_year_issued, 
    y = n
  )) + geom_point() + geom_line() + geom_vline(xintercept = 2020, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + # KH: felt important to have markers of the recession + pandemic
  labs(
    title = "District 3",
    y = "New Construction Permits", 
    x = "Year"
  )
```

### District 4
```{r}
districts |> 
  filter(council_district == "4") |>
  ggplot(aes(
    x = calendar_year_issued, 
    y = n
  )) + geom_point() + geom_line() + geom_vline(xintercept = 2020, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + # KH: felt important to have markers of the recession + pandemic
  labs(
    title = "District 4",
    y = "New Construction Permits", 
    x = "Year"
  )
```

### District 5
```{r}
districts |> 
  filter(council_district == "5") |>
  ggplot(aes(
    x = calendar_year_issued, 
    y = n
  )) + geom_point() + geom_line() + geom_vline(xintercept = 2020, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + # KH: felt important to have markers of the recession + pandemic
  labs(
    title = "District 5",
    y = "New Construction Permits", 
    x = "Year"
  )

# KH: They just run out of room?
```

### District 6
```{r}
districts |> 
  filter(council_district == "6") |>
  ggplot(aes(
    x = calendar_year_issued, 
    y = n
  )) + geom_point() + geom_line() + geom_vline(xintercept = 2020, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + # KH: felt important to have markers of the recession + pandemic
  labs(
    title = "District 6",
    y = "New Construction Permits", 
    x = "Year"
  )

# KH: Mostly single family homes is this area? Lots of traction before the recession, crawling back up and then drops at the Pandemic?
```

### District 7
```{r}
districts |> 
  filter(council_district == "7") |>
  ggplot(aes(
    x = calendar_year_issued, 
    y = n
  )) + geom_point() + geom_line() + geom_vline(xintercept = 2020, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + # KH: felt important to have markers of the recession + pandemic
  labs(
    title = "District 7",
    y = "New Construction Permits", 
    x = "Year"
  )
```
### District 8
```{r}
districts |> 
  filter(council_district == "8") |>
  ggplot(aes(
    x = calendar_year_issued, 
    y = n
  )) + geom_point() + geom_line() + geom_vline(xintercept = 2020, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + # KH: felt important to have markers of the recession + pandemic
  labs(
    title = "District 8",
    y = "New Construction Permits", 
    x = "Year"
  )
```

### District 9
```{r}
districts |> 
  filter(council_district == "9") |>
  ggplot(aes(
    x = calendar_year_issued, 
    y = n
  )) + geom_point() + geom_line() + geom_vline(xintercept = 2020, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + # KH: felt important to have markers of the recession + pandemic
  geom_vline(xintercept = 2004, color = "purple", linetype ="dashed") + # KH: This is when the University Neighborhood Overlay was introduced, which wanted to foster more higher density growth in West Campus
  labs(
    title = "District 9",
    y = "New Construction Permits", 
    x = "Year"
  )

#KH: For this section, I would also want to look at renovations? Assuming that lower density structures would need to be demlioshed in order to place higher density builds would mean we need new construction permits, right?

```

### District 10
```{r}
districts |> 
  filter(council_district == "10") |>
  ggplot(aes(
    x = calendar_year_issued, 
    y = n
  )) + geom_point() + geom_line() + geom_vline(xintercept = 2020, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2008, color = "red", linetype = "dashed") + # KH: felt important to have markers of the recession + pandemic
  labs(
    title = "District 10",
    y = "New Construction Permits", 
    x = "Year"
  )
```